Calibrating MEM seeding heuristics

### 2018-05-22 MEM.tex
Rewrote the theory without state ~ (seems to give identical results).

The weighted generating function of the transition from $\Downarrow$ to
$(\uparrow)_i$ is
\begin{equation}
u_i(z) = (\xi_{i-1,N}-\xi_{i,N})(qz)^i.
\end{equation}

@@

The $\downarrow$ symbol can be followed by all the terminators. We need to
start with the case of the $\sim$ terminator, because it will be required
for the other ones. Referring to the decision tree of
Fig.~\ref{fig:decision_tree}, we see that the $\sim$ symbol occurs in an
error-free stretch, if all the strictly masking threads fail before the
potentially masking threads. In section~\ref{sec:trans_from_sim} it will
be important to distinguish different cases based on the value of
$\varphi_0$ at the position of the $\sim$ symbol, so as in
section~\ref{sec:trans_from_up}, we introduce the states $(\sim)_j$ $(1
\leq j \leq \gamma-1)$ such that $\varphi_0(n) = j$ if the nucleotide at
position $n$ is $(\sim)_j$. In this case, the value of $\varphi_0(n) = j$
is simply the position of the $\sim$ symbol relative to the preceding
$\downarrow$ symbol.

If there are $m = 0$ potentially masking threads, the next terminator
cannot be $\sim$. Assuming that there are $m > 0$ potentially masking
threads at the $\downarrow$ symbol, the probability that the have not all
faied at position $i$ is $\xi_{i,m}$. Similarly, the probability that the
last strictly masking thread fails at position $i$ is
$\xi_{i-1,N-m}-\xi_{i,N-m}$. So the probability that the next terminator
is $(\sim)_i$ is $q^j W_i$, where
\begin{eqnarray}
\label{eq:Si}
W_i &=& \sum_{m=0}^{N-1} Q_m \cdot \xi_{i,m} \cdot
  (\xi_{i-1,N-m}-\xi_{i,N-m}) \notag \\
  &=& \frac{\alpha_i^N-\alpha_{i-1}^N-\gamma_i^N+\delta_{i-1}^N}
{1-(1-\mu/3)^N} \text{, where}
\end{eqnarray}

\begin{eqnarray}
\gamma_i &=& 1-(1-\mu)^i \\
\delta_i &=& 1-(1-\mu+\mu^2/3)(1-\mu)^i.
\end{eqnarray}

\begin{remark}
The term $\alpha_i^N-\alpha_i^N$ is the probability that the last dupicate
thread fails in one error followed by $i$ error-free nucleotides. The term
$\gamma_i^N - \delta_{i-1}^N$ is the probability that all threads fail in
$i$ error-free nucleotides...
\end{remark}

We have assumed $m>0$, but the sum above can start at $m=0$ because
$\xi_{i,0}=0$. The weighted generating function of the transition from
$\downarrow$ to $(\sim)_i$ is
\begin{equation}
w_i(z) = \frac{\alpha_i^N-\alpha_{i-1}^N-
  \gamma_i^N+\delta_{i-1}^N}{1-(1-\mu/3)^N} \cdot (qz)^i.
\end{equation}

\begin{remark}
When $N = 1$, $V_i = 0$ and the states $(\sim)_i$ are inaccessible.
The symbol $\sim$ can exist only if there is more than one duplicate.
\end{remark}

Referring again to the decision of Fig.~\ref{fig:decision_tree}, we see
that the $\uparrow$ symbol occurs in an error-free stretch, if all the
potentially masking threads fail before or at the same nucleotide as the
strictly masking threads. Assuming that there are $m \geq 0$ potentially
masking threads at the $\downarrow$ symbol, the probability that they have
all failed at position $i$ is $1-\xi_{i,m}$. As we have seen above, the
probability that the last strictly masking thread fails at position $j$ is
$\xi_{i-1,N-m}-\xi_{i,N-m}$. So the probability that the next terminator
is the $\uparrow$ symbol at position $i$ is $q^i W_i$, where
\begin{eqnarray}
  V_i &=& \sum_{m=0}^{N-1} Q_m \cdot (1-\xi_{i,m}) \cdot
  (\xi_{i-1,N-m}-\xi_{i,N-m}) \notag \\
    &=& \frac{\gamma_i^N-\delta_{i-1}^N}{1-(1-\mu/3)^N}.
\end{eqnarray}

The weighted generating function of the transition from $\downarrow$ to
$(\uparrow)_i$ is
\begin{equation}
v_i(z) = \frac{\gamma_i^N- \delta_{i-1}^N}{1-(1-\mu/3)^N} \cdot (qz)^i.
\end{equation}

The probability that there is neither $\sim$ nor $\uparrow$ symbol is just
the probability that at least one stricty masking thread is still alive
\begin{equation}
\sum_{m=0}^{N-1}Q_m \cdot \xi_{i,N-m} = \frac{1-\alpha_i^N}{1-(1-\mu/3)^N}
\end{equation}

So the weighted generating function of the transition from $\downarrow$ to
$\Downarrow$ is
\begin{equation}
B^\Downarrow(z) = \omega z\sum_{i=0}^\infty
\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
\end{equation}

If the next terminator is $\downarrow$, we again have to distinguish two
cases. In the example below there is no MEM seed, but the nature of the
terminator depends on $\gamma$. If $\gamma < 10$, the state $(\sim)_9$
does not exist, so there is a single segment of size 12, terminated by
$\downarrow$. If $\gamma \geq 10$, the state $(\sim)_9$ exists and there
is a segment of size 9 terminated by $\sim$, followed by a segment of size
3 terminated by $\downarrow$.
\begin{inset}
\includegraphics[scale=.9]{down_to_down_case_1.pdf}
\end{inset}

So if the two $\downarrow$ symbols are separated by $\gamma-1$ or fewer
nucleotides, there is no MEM seed and we must only rule out the presence
of other terminators in the segment (in this case $\sim$ and $\uparrow$).
With the probability computed above that there are neither $\sim$ nor
$\uparrow$ symbols, the weighted generating function is
\begin{equation}
\tilde{\omega} z\sum_{i=0}^{\gamma-1}
\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
\end{equation}

If the two $\downarrow$ symbols are separated by $\gamma$ or more
nucleotides, there could be a shared MEM seed, so we must enforce the
condition that the main thread is masked. In addition, we must also rule
out the presence of the $\sim$ terminator among the first $\gamma-1$
nucleotides (but not among the remaining nucleotides). Finally, we must
rule out the presence of the $\uparrow$ terminator throughout the segment.
This is satisfied if and only if $i.$ a strictly masking thread survives
until the penultimate nucleotide of the fragment or $ii.$ all the strictly
masking threads fail before the penultimate nucleotide but not before the
$\gamma$-th nucleotide, and a potentially masking thread survives until
the last nucleotide of the fragment.

If there are $m$ potentially masking threads and $N-m$ strictly masking
threads, the respective probabilities of conditions $i.$ and $ii.$ are
$\xi_{i-1,N-m} \cdot (1-(1-\mu/3)^N)$ and $(\xi_{i-1,N-m}
-\xi_{\gamma-1,N-m})(1-\alpha_{i-1}^m)$. Since the conditions are
are mutually exclusive, the full probability is
\begin{equation}
\sum_{m=0}^{N-1}Q_m \bigg(\xi_{i-1,N-m} \cdot (1-(1-\mu/3)^N) +
(\xi_{\gamma-1,N-m}-\xi_{i-1,N-m})(1-\alpha_{i-1}^m) \bigg).
\end{equation}

The first term is equal to $1-\alpha_{i-1}^N$, the second is equal to
\begin{equation}
\frac{\alpha_{i-1}^N - \alpha_{\gamma-1}^N -
  \zeta_{i-1,i-1}^N + \zeta_{\gamma-1,i-1}^N}{1-(1-\mu/3)^N}.
\end{equation}
where
\begin{equation}
\zeta_{j,i} = 1-(1-\mu)^j\mu/3-(1-\mu)^i\mu/3(1-\mu/3)
\end{equation}

Finally, the weighted generating function of the transitions from
$\downarrow$ to $\downarrow$ is

\begin{equation*}
\begin{split}
B^\downarrow(z) = \tilde{\omega} z \sum_{i=0}^{\gamma-1}
& \frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i + \\
pz\sum_{i=\gamma}^\infty & \left( 1-\alpha_i^N + 
\frac{\alpha_i^N - \alpha_{\gamma-1}^N -
  \zeta_{i,i}^N + \zeta_{\gamma-1,i}^N}{1-(1-\mu/3)^N} \right)
(qz)^i
\end{split}
\end{equation*}

\subsection{Transitions from $\sim$}
\label{sec:trans_from_sim}

At a $\sim$ symbol, the main thread becomes co-dominant, together with $1$
to $N-1$ alternative threads (if there were $N$, the symbol would be
$\Downarrow$). As in section~\ref{sec:trans_from_Down}, the next
terminator symbol depends on whether all the alternative co-dominant
threads fail before the main thread. If this is the case, the main thread
will become dominant ($\uparrow$ symbol), otherwise we again need to
distinguish the case that the read is reset ($\Downarrow$ symbol) from the
case that the main thread becomes dominated ($\downarrow$ symbol).

An important difficulty is that the $\sim$ symbol does not inform us about
the number of surviving threads (except when $N = 1$). So we need to
marginalize the probabilities of interest over this number. The first
thing to note is that a $\sim$ terminator can only follow a $\downarrow$
terminator. Indeed, after a $\Downarrow$ symbol the threads are already
co-dominant, and after a $\uparrow$ symbol the alternative threads cannot
`catch up' with the main thread until a sequencing error occurs,
\textit{i.e.} until a $\Downarrow$ symbol or a $\downarrow$ symbol is
inserted. The number of surviving threads depends on the position of the
$\sim$ symbol relative to the preceding $\downarrow$ symbol, so we need
to treat the states $(\sim)_j$ separately.

We first establish an equality that will turn out to be useful in the
subsequent calculations.

\begin{lemma}
For any real numbers $x$ and $y$,
\begin{eqnarray*}
\sum_{n=m}^N {N \choose n} {n \choose m} x^n y^{N-n}
&=& x^m \sum_{r=0}^{N-m} {N \choose m+r} {m+r \choose m}a^r y^{N-m-r} \\
&=& {N \choose m} x^m \sum_{r=0}^{N-m} {N-m \choose r}x^r y^{N-m-r} \\
&=& {N \choose m} x^m (x+y)^{N-m}.
\end{eqnarray*}

The equality above implies
\begin{equation}
\label{eq:double_binom}
\sum_{n=m}^{N-1} {N \choose n} {n \choose m} x^n y^{N-n} =
{N \choose m} x^m \Big[ (x+y)^{N-m} - x^{N-m} \Big].
\end{equation}
\end{lemma}

The probability that there are $m>1$ potentially masking threads upon
entering state $(\sim)_j$ is the probability that there were $n \geq m$
potentially masking threads at the $\downarrow$ symbol -- probability
$Q_n$ --, that exactly $m$ have survived until position $j$ -- probability
${n \choose m} (1-\mu)^{jm} (1-(1-\mu)^j)^{n-m}$ -- and that the last of
the $N-n$ strictly masking threads fails at position $j$ -- probability
$\xi_{j-1,N-n}-\xi_{j,N-n}$. Summing over $n = m, \ldots, N-1$, the
probability is seen to be
\begin{equation}
\label{eq:miscprob}
P_{m,j} = \sum_{n=m}^{N-1} Q_n \cdot {n \choose m} (1-\mu)^{jm}
(1-(1-\mu)^j)^{n-m} \cdot (\xi_{j-1,N-n}-\xi_{j,N-n}).
\end{equation}

We write $\xi_{j-1,N-n}-\xi_{j,N-n}$ as $(1-\xi_{j,N-n}) -
(1-\xi_{j-1,N-n})$ and split the sum of equation (\ref{eq:miscprob}) into
two parts. The first term is equal to
\begin{equation*}
\begin{split}
\sum_{n=m}^{N-1} & Q_n \cdot {n \choose m} (1-\mu)^{jm} (1-(1-\mu)^j)^{n-m}
\cdot (1 - \xi_{j,N-n}) = \\
& \frac{(1-\mu)^{jm} (1-(1-\mu)^j)^{N-m}}{1-(1-\mu/3)^N}
\sum_{n=m}^{N-1} {N \choose n} {n \choose m} (1-\mu/3)^n (\mu/3)^{N-n}.
\end{split}
\end{equation*}

Using formula (\ref{eq:double_binom}) with $x=1-\mu/3$ and $y=\mu/3$, we
see that this term is equal to
\begin{equation*}
\frac{(1-\mu)^{jm} (1-\mu/3)^m}{1-(1-\mu/3)^N} {N \choose m} 
\bigg( \gamma_j^{N-m}
-\big[ (1-(1-\mu)^j)(1-\mu/3)\big]^{N-m} \bigg).
\end{equation*}

The second term of equation (\ref{eq:miscprob}) is equal to
\begin{multline*}
\sum_{n=m}^{N-1} Q_n \cdot {n \choose m} (1-\mu)^{jm} (1-(1-\mu)^j)^{n-m}
\cdot (1 - \xi_{j-1,N-n}) = \\
\frac{1}{1-(1-\mu/3)^N}
\left(\frac{(1-\mu)^j}{1-(1-\mu)^j}\right)^m
\; \sum_{n=m}^{N-1} {N \choose n} {n \choose m} \times \\
(1-\mu/3)^n (\mu/3)^{N-n}
(1-(1-\mu)^j)^n (1-(1-\mu)^{j-1})^{N-n}.
\end{multline*}

Formula (\ref{eq:double_binom}) with $x=(1-(1-\mu)^j)(1-\mu/3)$ and
$y = (1-(1-\mu)^{j-1})\mu/3$ yields
\begin{equation*}
\frac{(1-\mu)^{jm} (1-\mu/3)^m}{1-(1-\mu/3)^N} {N \choose m}
\bigg( \delta_{j-1}^{N-m}
-\big[ (1-(1-\mu)^j)(1-\mu/3)\big]^{N-m} \bigg).
\end{equation*}

Computing the difference between the terms, we see that the probability
that there are $m$ potentially masking threads upon entering state
$(\sim)_j$ is
\begin{equation*}
P_{m,j} = \frac{(1-\mu)^{jm} (1-\mu/3)^m}{1-(1-\mu/3)^N} {N \choose m} 
\big( \gamma_j^{N-m} - \delta_{j-1}^{N-m} \big).
\end{equation*}

The sum of the terms above for $m = 1, \ldots, N-1$ is $W_j$ from equation
(\ref{eq:Si}), which is the probability of the $(\sim)_j$ terminator.

Now, referring to the decision tree of Fig.~\ref{fig:decision_tree}, we
see that if the $m$ potentially masking threads fail before the main
thread, the next terminator will be the $\uparrow$ symbol, otherwise it
will be either the $\Downarrow$ symbol or the $\downarrow$ symbol.

The probability that the last of the $m$ potentially masking threads fails
at position $i$ after entering state $(\sim)_j$ is
$\xi_{i-1,m}-\xi_{i,m}$. Dividing by $W_j$ and summing over $m = 1,
\ldots, N-1$, we find that the weighted generating function of the
transition from $(\sim)_j$ to $(\uparrow)_{j+i}$ is
\begin{eqnarray}
y_{j,i}(z) &=& \sum_{m=1}^{N-1} \frac{P_{m,j}}{W_j}
\big( \xi_{i-1,m}-\xi_{i,m} \big) (qz)^i \notag \\
&=& \frac{\beta_{j,i+j}^N - \beta_{j,i+j-1}^N
-\beta_{j-1,i+j}^N + \beta_{j-1,i+j-1}^N}
  {\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i,
\end{eqnarray}
where $\alpha_j$, $\delta_j$, $\gamma_j$ are as above, and where
\begin{equation}
\beta_{j,i} = 1 - (1-\mu)^j\mu/3 - (1-\mu)^i(1-\mu/3).
\end{equation}

The probability that the next terminator is a $\Downarrow$ symbol or a
$\downarrow$ symbol at position $i$ is the probability that at least one
of the potentially masking threads survives until position $i-1$ --
probability $\xi_{i-1,m}$ -- and that the next nucleotide is a sequencing
error -- probability $p$. Dividing by $W_j$ and summing over $m = 1,
\ldots, N-1$, we find that the weighted generating function of the
transtion from $(\sim)_j$ to
$\Downarrow$ is
\begin{eqnarray}
D^\Downarrow_j(z) &=& \omega z \sum_{i=0}^{\gamma-j-2} \sum_{m=1}^{N-1}
  \frac{P_{m,j}}{W_j} \xi_{i,m} \cdot (qz)^i \notag \\
  &=& \omega z \sum_{i=0}^{\gamma-j-2}
\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j}^N
  +\beta_{j-1,i+j}^N }
{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i.
\end{eqnarray}

With the same notations, the weighted generating function from $(\sim)_j$
to $\downarrow$ is seen to be 
\begin{eqnarray}
D^\downarrow_j(z) = \tilde{\omega} z
  \sum_{i=0}^{\gamma-j-2} \sum_{m=1}^{N-1}
  \frac{P_{m,j}}{W_j} \xi_{i,m} \cdot (qz)^i +
  pz \sum_{i=\gamma-j-1}^\infty \sum_{m=1}^{N-1}
  \frac{P_{m,j}}{W_j} (1-\alpha_i^m) \cdot (qz)^i \notag \\
\tilde{\omega} z \sum_{i=0}^{\gamma-j-2}
\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j}^N
  +\beta_{j-1,i+j}^N }
{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i + \notag \\
pz \sum_{i=\gamma-j-1}^\infty
\frac{ \alpha_j^N-\alpha_{j-1}^N-\zeta_{j,i+j}^N
  +\zeta_{j-1,i+j}^N }
{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i + \notag \\
\end{eqnarray}

@@

At the beginning of the read, the match lengths of all the threads are
$0$. This is exactly the same as on a $\Downarrow$ symbol. We can thus
consider that the read starts in the $\Downarrow$ state, which is
equivalent to adding an edge labelled with the empty object $\varepsilon$
between the head vertex and the $\Downarrow$ vertex. The head vector is
thus
\begin{equation*}
H(z) = 
\begin{blockarray}{cccccccc}
   \;\; \DN & \dn &\eq{1} & \ldots & \eq{\gamma-1} &
    \up{1} & \ldots & \up{\gamma-1} \\
\begin{block}{[cccccccc]}
\;\; 1 & 0 & 0 & \ldots & 0 & 0 & \ldots & 0 \\
\end{block}
\end{blockarray}.
\end{equation*}
The tail vector contains the weighted generating functions of
non-terminated stretches of $\square$ symbols. Referring to the decision
tree of Fig.~\ref{fig:decision_tree}, we see that the probability of
occurrence of $\square$ symbols depends on whether there are potentially
masking and / or strictly masking threads. All the symbols except
$\downarrow$ impose a maximum on the number of $\square$ symbols at the
end of the reads.

For the $\downarrow$ symbol, the weighted generating function is
\begin{equation}
T^\downarrow(z) = \sum_{i=0}^\infty
\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
\end{equation}

@@

For the $(\sim)_j$ state, the weighted generating function is
\begin{equation}
T^\sim_j(z) = \sum_{i=0}^{\gamma-j-1}
\frac{ \alpha_j^N-\alpha_{j-1}^N-\beta_{j,i+j}^N +\beta_{j-1,i+j}^N }
{\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i.
\end{equation}

In summary,
\begin{equation*}
T(z) = 
\begin{blockarray}{cc}
\begin{block}{c[c]}
\DN & T^\Downarrow(z) \\
\dn & T^\downarrow(z) \\
\up{\gamma-1} & T^\uparrow_{\gamma-1}(z) \\
\vdots & \vdots \\
\up{1} & \; T^\uparrow_1(z) \; \\
\eq{\gamma-1} & T^\sim_{\gamma-1}(z) \\
\vdots & \vdots \\
\eq{1} & T^\sim_1(z) \\
\end{block}
\end{blockarray}.
\end{equation*}

@@

\begin{equation*}
\begin{blockarray}{ccccccccc}
   & \DN & \dn &\up{\gamma-1} & \ldots & \up{1} &
    \eq{\gamma-1} & \ldots & \eq{1} \\
\begin{block}{c[cccccccc]}
\DN & A_\gamma(z) & \tilde{A}_\infty(z) & u_{\gamma-1}(z) & \ldots &
    u_1(z) & 0 & \ldots & 0 \\
\dn & B_\infty(z) & \tilde{B}_\infty(z) & v_{\gamma-1}(z) & \ldots &
    v_1(z) & w_{\gamma-1}(z) & \ldots & w_1(z) \\
\up{\gamma-1} & C_1(z) & \tilde{C}_1(z) & 0 & 
    \ldots & 0 & 0 & \ldots & 0 \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
    \ddots & \vdots \\
\up{1} & C_{\gamma-1}(z) & \tilde{C}_{\gamma-1}(z) & 0 & \ldots & 0 & 0 &
    \ldots & 0 \\
\eq{\gamma-1} & D_1(z) & \tilde{D}_1(z) & 0 & \ldots & 0 & 0 &
    \ldots & 0 \\
\eq{\gamma-2} & D_2(z) & \tilde{D}_2(z) & y_{\gamma-2,1}(z) &
    \ldots & 0 & 0 & \ldots & 0 \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
    \ddots & \vdots \\
\eq{2} & D_{\gamma-2}(z) & \tilde{D}_{\gamma-2}(z) &
    y_{2,\gamma-3}(z) & \ldots & 0 & 0 & \ldots & 0 \\
\eq{1} & D_{\gamma-1}(z) & \tilde{D}_{\gamma-1}(z) &
    y_{1,\gamma-2}(z) & \ldots & 0 & 0 & \ldots & 0 \\
\end{block}
\end{blockarray}
\end{equation*}

In the case $N = 1$, reads never contain the symbol $\sim$ and the matrix
is
\begin{equation*}
\begin{blockarray}{cccccc}
   & \DN & \dn & \up{1} & \ldots & \up{\gamma-1} \\
\begin{block}{c[ccccc]}
\DN & A_\gamma(z) & \tilde{A}_\infty(z) & u_1(z)
    & \ldots & u_{\gamma-1}(z) \\
\dn & B_\infty(z) & \tilde{B}_\infty(z) & v_1(z)
    & \ldots & v_{\gamma-1}(z) \\
\up{1} & D_{\gamma-1}(z) & \tilde{D}_{\gamma-1}(z) & 0 &
    \ldots & 0 \\
\up{2} & D_{\gamma-2}(z) & \tilde{D}_{\gamma-2}(z) & 0 &
    \ldots & 0  \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots \\
\up{\gamma-1} & D_1(z) & \tilde{D}_1(z) & 0 & \ldots & 0 \\
\end{block}
\end{blockarray}
\end{equation*}

In the case $N = 0$, the symbol $\downarrow$ also disappears. Some
polynomials have a particularly simple form in this case, and the
body of the transfer matrix is
\begin{equation*}
\begin{blockarray}{ccc}
   & \DN & \up{1} \\
\begin{block}{c[cc]}
\DN    & pz              & qz \\
\up{1} & D_{\gamma-1}(z) & 0  \\
\end{block}
\end{blockarray}
\end{equation*}

>>>>

The weighted generating function of the transition from $\Downarrow$ to
$(\uparrow)_i$ is
\begin{equation}
\label{eq:u}
u_i(z) = (\xi_{i-1,N}-\xi_{i,N})(qz)^i.
\end{equation}

@@

The probability that the next terminator is the $\uparrow$ symbol at
position $i$ is the probability that the last of the $N$ duplicate threads
fails at position $i$ is $\xi_{i-1,N-m}-\xi_{i,N-m}$ so the weighted
generating function of the transition from $\downarrow$ to $(\uparrow)_i$
is
\begin{equation}
\tag{\ref{eq:u}}
u_i(z) = (\xi_{i-1,N}-\xi_{i,N}) \cdot (qz)^i.
\end{equation}

To compute the weighted generating function of the transition from
$\downarrow$ to $\Downarrow$, we need to separate two cases. In the first,
the error symbols are separated by $\gamma-1$ or fewer error-free
nucleotides. In this case, there can be no MEM seed and we only have to
exclude the $\uparrow$ terminator. We obtain
\begin{equation}
\omega z \sum_{i=0}^{\gamma-1}\xi_{i,N} \cdot (qz)^i.
\end{equation}

In the case that $\downarrow$ and $\Downarrow$ are separated by $\gamma$
or more error-free nucleotides, we must ensure that at least one strictly
masking thread survives until the penultimate nucleotide of the segment.
This probability is equal to
\begin{equation}
\sum_{m=0}^{N-1}Q_m \cdot \xi_{i,N-m} =
\frac{1-\alpha_i^N}{1-(1-\mu/3)^N}.
\end{equation}

\begin{remark}
A strictly masking thread survives until the $\Downarrow$ symbol if and
only if one of the $N$ duplicate threads reverse-survives from
the penultimate nucleotide of the segment down to and including the
$\downarrow$ symbol.
\end{remark}

In sum, the weighted generating function of the transition from
$\downarrow$ to $\Downarrow$ is
\begin{equation}
B^\Downarrow(z) = 
\omega z \sum_{i=0}^{\gamma-1}\xi_{i,N} \cdot (qz)^i +
\omega z\sum_{i=\gamma}^\infty \frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
\end{equation}

If the next terminator is $\downarrow$, we again have to distinguish two
cases. If the two $\downarrow$ symbols are separated by $\gamma-1$ or
fewer nucleotides, there is no MEM seed and we must only rule out the
presence of the $\uparrow$ terminator. We obtain
\begin{equation}
\tilde{\omega} z\sum_{i=0}^{\gamma-1}
\xi_{i,N} \cdot (qz)^i.
\end{equation}

If the two $\downarrow$ symbols are separated by $\gamma$ or more
nucleotides, there could be a shared MEM seed, so we must enforce the
condition that the main thread is masked. This is satisfied if and only if
one of the following conditions holds:
\begin{enumerate}
\item A strictly masking thread survives until the penultimate
nucleotide of the segment.
\item All the strictly masking threads fail before the penultimate
nucleotide but a potentially masking thread survives until the last
nucleotide of the segment.
\end{enumerate}

Assuming that there are $m \geq 0$ potentially masking threads, the
probability of condition $1.$ is $\xi_{i-1,N-m}$ and that of condition 2.
is $(1-\xi_{i-1,N-m}) (1-\alpha_{i-1}^m)$. Summing of $m = 0, \ldots, N-1$
on obtains
\begin{equation}
(1-\mu/3)\sum_{m=0}^{N-1}Q_m \cdot \xi_{i-1,N-m} =
1-\alpha_{i-1}^N \text{, and}
\end{equation}
\begin{equation}
\sum_{m=0}^{N-1}Q_m (1-\xi_{i-1,N-m}) (1-\alpha_{i-1}^m) =
\frac{\alpha_{i-1}^N - \alpha_0^N -
  \zeta_{i-1,i-1}^N + \zeta_{0,i-1}^N}{1-(1-\mu/3)^N},
\end{equation}
where
\begin{equation}
\zeta_{j,i} = 1-(1-\mu)^j\mu/3-(1-\mu)^i\mu/3(1-\mu/3)
\end{equation}

Finally, the weighted generating function of the transitions from
$\downarrow$ to $\downarrow$ is
\begin{equation*}
B^\downarrow(z) = \tilde{\omega}z \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot
(qz)^i + pz \sum_{i=\gamma}^\infty
\left( 1-\alpha_i^N +
\frac{\alpha_i^N - \alpha_0^N - \zeta_{i,i}^N +
\zeta_{0,i}^N}{1-(1-\mu/3)^N} \right) (qz)^i
\end{equation*}

\begin{remark}
One can interpret $\zeta_{i,i}$ as the probability that
\end{remark}

@@

At the beginning of the read, the match lengths of all the threads are
$0$. This is exactly the same as on a $\Downarrow$ symbol. We can thus
consider that the read starts in the $\Downarrow$ state, which is
equivalent to adding an edge labelled with the empty object $\varepsilon$
between the head vertex and the $\Downarrow$ vertex. The head vector is
thus
\begin{equation*}
H(z) = 
\begin{blockarray}{ccccc}
   \;\; \DN & \dn & \up{1} & \ldots & \up{\gamma-1} \\
\begin{block}{[ccccc]}
\;\; 1 & 0 & 0 & \ldots & 0 \\
\end{block}
\end{blockarray}.
\end{equation*}
The tail vector contains the weighted generating functions of
non-terminated stretches of $\square$ symbols. Referring to the decision
tree of Fig.~\ref{fig:decision_tree}, we see that the probability of
occurrence of $\square$ symbols depends on whether there are potentially
masking and / or strictly masking threads. All the symbols except
$\downarrow$ impose a maximum on the number of $\square$ symbols at the
end of the reads.

For the $\downarrow$ symbol, the weighted generating function is
\begin{equation}
T^\downarrow(z) = 
\sum_{i=0}^{\gamma-1}\xi_{i,N} \cdot (qz)^i +
\sum_{i=\gamma}^\infty \frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
\end{equation}

@@

In summary,
\begin{equation*}
T(z) = 
\begin{blockarray}{cc}
\begin{block}{c[c]}
\DN & T^\Downarrow(z) \\
\dn & T^\downarrow(z) \\
\up{\gamma-1} & T^\uparrow_{\gamma-1}(z) \\
\vdots & \vdots \\
\up{1} & \; T^\uparrow_1(z) \; \\
\end{block}
\end{blockarray}.
\end{equation*}

@@

\begin{equation*}
\begin{blockarray}{cccccc}
   & \DN & \dn & \up{1} & \ldots & \up{\gamma-1} \\
\begin{block}{c[ccccc]}
\DN & A^\Downarrow(z) & A^\downarrow(z) & u_1(z)
    & \ldots & u_{\gamma-1}(z) \\
\dn & B^\Downarrow(z) & B^\downarrow(z) & u_1(z)
    & \ldots & u_{\gamma-1}(z) \\
\up{1} & C_{\gamma-1}(z) & \tilde{C}_{\gamma-1}(z) & 0 &
    \ldots & 0 \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots \\
\up{\gamma-1} & C_1(z) & \tilde{C}_1(z) & 0 & \ldots & 0 \\
\end{block}
\end{blockarray}
\end{equation*}

In the case $N = 0$, the symbol $\downarrow$ also disappears. Some
polynomials have a particularly simple form in this case, and the
body of the transfer matrix is
\begin{equation*}
\begin{blockarray}{ccc}
   & \DN & \up{1} \\
\begin{block}{c[cc]}
\DN    & pz              & qz \\
\up{1} & C_{\gamma-1}(z) & 0  \\
\end{block}
\end{blockarray}
\end{equation*}
